---
title: " Elliptic Fourier Analysis explained and based on Grapevine seeds/pips"
subtitle:  "INRA paleogenetics grapevine project GMM pipeline (R)"
format: gfm
execute:
    warning: false
---

## 1. Mathematical background
A closed 2D contour (seed outline) is treated as a periodic function of arc length length t ∈ [0, T), T = perimeter. x(t) and y(t) are each expanded as separated Fourier series (**Kuhl & Giardina 1982**). Truncating at harmonic N reconstruct the outline as a sum of N ellipses traced at increasing frequency. Harminoc 1 is a single ellipse, harmonic 2 as second ellipse spinning twice as fast, etc...

For a digitized closed contour of K points, with Δx_p, Δy_p, Δt_p the per-segment increments and cumulative arc length t_p:

```
a_n = (T / 2n²π²) Σ (Δx_p/Δt_p)[cos(2nπt_p/T) − cos(2nπt_{p−1}/T)]
b_n = (T / 2n²π²) Σ (Δx_p/Δt_p)[sin(2nπt_p/T) − sin(2nπt_{p−1}/T)]
c_n = (T / 2n²π²) Σ (Δy_p/Δt_p)[cos(2nπt_p/T) − cos(2nπt_{p−1}/T)]
d_n = (T / 2n²π²) Σ (Δy_p/Δt_p)[sin(2nπt_p/T) − sin(2nπt_{p−1}/T)]

```
Reconstruction: x(t) = A₀ + Σ[a_n cos(2nπt/T) + b_n sin(2nπt/T)], y(t) analogous with c_n, d_n.

**Normalization** (removes starting point, rotation, size, tracing direction) is done via the first-harmonic ellipse: scale to its semi-major axis onto the x-axis, phase-shift the start point to its tip. After normalization, hamonic 1 carries no shape information; analysis uses harmonic 2...(n). Harmonic count N is chosen from cumulative Fourier power (Σ power_i / Σ_total), typically the smallest N reaching ~99%.

The normalized coefficient vector (4x(N-1) values per specimen) is the shape descriptor. Stacked across specimens -> PCA -> morphospace used to separate morphotypes (wild vs domesticated, cultivar groups, chronological/geographic strata).

## 2. Environment

```{r}
#| eval: false
install.packages("Momocs")
library(Momocs)
```

If the install doesnt work, follow the next install process,which is more accessible via windows.

```{r}
install.packages("remotes")
remotes::install_github("MomX/Momocs")
```

Momocs implements this exact algorithm (`efourier()`), plus PCA/MANOVA/LDA on the resulting coefficients, with a consistent grammar throughout (Bonhomme, Picq, Gaucherel & Claude 2014, *J. Stat. Softw.* 56(13)).

## 3. Input data requirements
`import_jpg()` expects **binary silhouette images**: the seed as solid black pixels on a white background, one specimen per image, roughly centered.Raw archaeobotanical photographs