---
title: " Elliptic Fourier Analysis explained and based on Grapevine seeds/pips"
subtitle:  "INRA paleogenetics grapevine project GMM pipeline (R)"
format: gfm
execute:
    warning: false
---

## 1. Mathematical background
A closed 2D contour (seed outline) is treated as a periodic function of arc length length t ∈ [0, T), T = perimeter. x(t) and y(t) are each expanded as separated Fourier series (**Kuhl & Giardina 1982**). Truncating at harmonic N reconstruct the outline as a sum of N ellipses traced at increasing frequency. Harminoc 1 is a single ellipse, harmonic 2 as second ellipse spinning twice as fast, etc...

For a digitized closed contour of K points, with Δx_p, Δy_p, Δt_p the per-segment increments and cumulative arc length t_p:

```
a_n = (T / 2n²π²) Σ (Δx_p/Δt_p)[cos(2nπt_p/T) − cos(2nπt_{p−1}/T)]
b_n = (T / 2n²π²) Σ (Δx_p/Δt_p)[sin(2nπt_p/T) − sin(2nπt_{p−1}/T)]
c_n = (T / 2n²π²) Σ (Δy_p/Δt_p)[cos(2nπt_p/T) − cos(2nπt_{p−1}/T)]
d_n = (T / 2n²π²) Σ (Δy_p/Δt_p)[sin(2nπt_p/T) − sin(2nπt_{p−1}/T)]

```
Reconstruction: x(t) = A₀ + Σ[a_n cos(2nπt/T) + b_n sin(2nπt/T)], y(t) analogous with c_n, d_n.

**Normalization** (removes starting point, rotation, size, tracing direction) is done via the first-harmonic ellipse: scale to its semi-major axis onto the x-axis, phase-shift the start point to its tip. After normalization, hamonic 1 carries no shape information; analysis uses harmonic 2...(n). Harmonic count N is chosen from cumulative Fourier power (Σ power_i / Σ_total), typically the smallest N reaching ~99%.

The normalized coefficient vector (4x(N-1) values per specimen) is the shape descriptor. Stacked across specimens -> PCA -> morphospace used to separate morphotypes (wild vs domesticated, cultivar groups, chronological/geographic strata).

## 2. Environment

```{r}
#| eval: false
install.packages("Momocs")
library(Momocs)
```

If the install doesnt work, follow the next install process,which is more accessible via windows.

```{r}
install.packages("remotes")
remotes::install_github("MomX/Momocs")
```

Momocs implements this exact algorithm (`efourier()`), plus PCA/MANOVA/LDA on the resulting coefficients, with a consistent grammar throughout (Bonhomme, Picq, Gaucherel & Claude 2014, *J. Stat. Softw.* 56(13)).

## 3. Input data requirements
`import_jpg()` expects **binary silhouette images**: the seed as solid black pixels on a white background, one specimen per image, roughly centered.Raw archaeobotanical photographs (RGB, multiple seeds per frame - see section 6)

If you already digitalized (x,y) outline coordinates (e.g. from Fiji or TPS file), skip `import_jpg()` and build the `Out` object directly from a list of coordinate matrices, or use `import_tps()`.

Keep a metadata table(`data.frame` or `.csv`) in the **same row order** as the image list: sampleID,site,chronological phase, preservationInfo (charred/waterlogged), and morphotype/domestication status where known. This becomes the `$fac` slot used for grouping in PCA/MONAVA/LDA, and is what you`ll later join against the aDNA sampleIDs.

## 4. Pipeline
```{r}
#| eval: false
library(Momocs)

# --- Import ---
jpg.paths <- list.files("data/seed_silhouettes", pattern "\\.jpg$", full.names = TRUE)
coo.list <- import_jpg(jpg.paths, threshold = 0.5)

meta <- read.csv("data/seed_metadata.csv") #file names are examples pls change them with real data
seeds <- Out(coo.list, fac = meta)  #check Out for exact arg name in the version

# --- Sanity check of the outlines ---
panel(seeds, names = TRUE)
seeds <-coo_check(seeds) # flags/handles malformed outlines (self-intersections, open contours)

# --- How many harmonics? ---
calibrate_harmonicpower_efourier(seeds, nb.h = 20)  #cumulative power table + plot
calibrate_reconstruction_efourier(seeds, id = 1, range 1:12) #visual check on one specimen

# --- Elliptic Fourier analysis ---
# norm = TRUE applies the standard first-harmonic-ellipse normalization(size/rotation/start point)
seeds.ef <- efourier(seeds, nb.h = 9, smooth.it = 0, norm = TRUE)

# --- PCA on the coefficient matrix ---
seeds.pca <- PCA(seed.ef)
plot(seeds.pca, "group")  # replace later "group" with your $fac column name
PCcontrib(seeds.pca, nax = 1:3) # Contains the 3 PCN axes

# --- Group separation ---
MANOVA(seeds.pca, "group")
seeds.lda <- LDA(seeds.ef, "group")
plot_CV(seeds.lda)

# --- Export for downstream analysis/linking to a DNA sampleIDs ---
write.csv(as_df(seeds.ef),"outputs/efa_coefficients.csv", row.names = FALSE)
write.csv(cbind(meta, seeds.pca$x), "output/efa_pca_scores.csv", row.names = FALSE)
```

# 5. Validation demo (no data required)

Momocs ships a demo `Out`dataset (`bot`, bottle outlines) that runs the full pipleine end-to-end - a fast way to confirm the setup works before pointing it at real seed data:

```{r}
#| eval: true
library(Momocs)
data(bot)
panel(bot, names = TRUE)

calibrate_harmonicpower_efourier(bot, nb.h = 12)

bot.ef <- efourier(bot, nb.h = 12, norm = TRUE)
bot.pca <- PCA(bot.ef)
plot(bot.pca, "type")
PCcontrib(bot.pca, nax = 1:3)
MANOVA(bot.pca, "type")
```

## 6. Adding to real archeaological seed photographs

Raw photos of charred/waterlogged pips are rarely one clean silhouette file. There are two pratical options you can follow:

- **`outlineR`** (`devtools::install_github("yesdavid/outlineR")`) -> It's a wrapper around Momocs + EBImage + imager built specifically for archaeological artefact photos: extracts mutiple outlines from a single image while preserving specimen names, optionally rescaling via a `.tps` file. Designed exactly for this use case.

- Manual route: threshold each photo to a binary mask (EBImage/imager/ImageJ),crop to one seed per file, then `import_jpeg()` as above.

## 7. Caveat: taphonomic distortion
Charring and waterlogging deform seed shape differently . Do not pool charred and waterlogged specimens as if shape-equivalent -> carry preservation mode as covariate/stratifying factor in `$fac`, and treat cross-preservation comparisons cautiously, consistent with practice in the refernce literature found on zotero.